How to select predictive models for causal inference?

### 2023-11-16 how_to_select_causal_models.tex
adapt images to two cols

%\documentclass[MYBIB,STIX1COL]{WileyNJD-v2}
\documentclass[10pt]{article}

% declared functions and additional packages
\usepackage{caption}
\usepackage{subcaption}
\usepackage{threeparttable}
\usepackage{times}
\usepackage{hyperref}
\def\mycitecolor{green!50!black}
\def\mylinkcolor{red!60!black}
\hypersetup{ colorlinks=true, linkcolor=\mylinkcolor, citecolor=\mycitecolor }
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{amsmath}
\usepackage{amssymb} %mathbb packages
\usepackage{siunitx}
\usepackage{algorithm, algorithmicx, algpseudocode}
\usepackage{multirow}
\usepackage[sorting=nyt,style=numeric, natbib=true,doi=false,isbn=false,url=false,eprint=false, maxnames=999, maxcitenames=2]{biblatex} 
\DeclareNameAlias{default}{last-first}
\usepackage{placeins}
% associated arxiv class
\usepackage{arxiv}
\usepackage{multicol}
\setlength{\columnsep}{1cm}
\usepackage{wrapfig}
%\usepackage{wraptable}
\usepackage{authblk}% for author affiliations
\usepackage[dvipsnames]{xcolor}

@@

\setlength{\marginparwidth}{1.5cm}
\setlength\textwidth{177.8mm}

@@

\newtheorem{definition}{Definition}
\newtheorem{condition}{Condition}
\newtheorem{property}{Property}
\newtheorem{remark}{Remark}
\newtheorem{notation}{Notation}
\newtheorem{proposition*}{Proposition}
\newtheorem{proposition}{Proposition}
\newtheorem{lemma}{Lemma}
\newtheorem{theorem}{Theorem}
\newtheorem{assumption}{Assumption}
\newtheorem{proof}{Proof}

@@

\affil[1]{Inria, Soda, Saclay, France}
\affil[2]{Mission Data, Haute Autorité de Santé, Saint-Denis, France}
\affil[*]{Corresponding author: matthieu.doutreligne@inria.fr}

@@

\maketitle

\begin{abstract}
    \textbf{Background:}
    As predictive models --\emph{eg} from machine learning-- give likely
    outcomes, they may be used to reason on the effect of an intervention,
    a causal-inference task. The increasing complexity of
    health data has opened the door to a plethora of models, but also the
    Pandora box of model selection: which of these models yield the most
    valid causal estimates? Causal model selection should control both outcome errors for each
    individual, treated or not treated, whereas only one outcome is observed.
    Theoretically, simple risks used in machine learning do not control causal
    effects when treated and non-treated population differ too much. \\
    \textbf{Methods:}
    More elaborate risks build proxies of the causal error using ``nuisance''
    re-weighting to compute it on the observed data. But does computing these
    nuisance adds noise to model selection? Drawing from an extensive empirical
    study based on a simulation and three semi-simulated datasets based on real
    covariates, we compare different metrics for causal model selection.\\
    \textbf{Results:} Among all metrics the classical mean squared error (ie.
    factual $\mu\mathrm{-risk}$) is worse and re-weighting it with propensity score
    ($\mu\mathrm{-risk}_{IPW}$) does not bring much improvements. The
    $R\text{-risk}$, which includes a model of mean outcome and propensity
    scores, leads to the best performances.\\
    \textbf{Conclusions:} We outline a good causal model-selection procedure
    using the so-called $R\text{-risk}$; using flexible estimators to compute
    the nuisance models on the train set; and splitting out 10\% of the data to
    compute risks.
\end{abstract}
\keywords{Model Selection, Treatment Effect, G-formula, Observational Study, Machine Learning, }
\clearpage

\begin{multicols}{2}

    \section{Background}\label{sec:intro}

@@

    \begin{figure}[b!]
        \begin{minipage}{.4\linewidth}
            \caption[The best predictor may not estimate best causal
                effects]{\textbf{Illustration}: a) a random-forest estimator
                with high performance for standard prediction (high $R^2$) but that
                yields poor causal estimates (large error between true effect $\tau$ and
                estimated $\hat{\tau}$), b) a linear estimator with smaller
                prediction performance leading to better causal estimation. \\[1ex]
                Selecting the estimator with the smallest error to the individual
                treatment effect $\mathbb{E}[(\tau(x) - \hat{\tau}(x))^2]$
                --the $\tau\text{-risk}$, def.\,\ref{def:tau_risk} -- would lead to
                the best causal estimates; however computing this error is not
                feasible: computing it requires access to unknown quantities:
                $\tau(x)$. \\[1ex]
                While the random forest fits the data better than the linear model, it
                gives worse causal inference because its error is very inhomogeneous between
                the treated and untreated. The $R^2$ score does not capture this
                inhomogeneity.
            }\label{fig:toy_example} \end{minipage}\hfill%
        \begin{minipage}{.577\linewidth}
            %\centering
            {\sffamily\small a) Random forest, good average prediction but bad
                causal inference}

            \hfill%
            \includegraphics[width=1\linewidth]{images/toy_random_forest_high_R2_high_tau_risk.pdf}%

            {\sffamily\small b) Linear model, worse average prediction but better causal inference}

            \hfill%
            \includegraphics[width=1\linewidth]{images/toy_tlinear_model_small_R2_small_tau_risk.pdf}%
        \end{minipage}%
    \end{figure}

    Using a predictor to reason on causal effects relies on contrasting the
    prediction of the outcome for a given individual with and without the
    treatment --as detailed in the Methods \autoref{sec:neyman_rubin}.
    %
    Given various predictors of the outcome, we are interested in
    selecting those that estimate best the treatment effect.
    %
    Standard predictive modeling or machine-learning practice selects the
    best predictor, \emph{ie} the one that minimizes the expected error.
    However, the best predictor may not be the best model to reason about
    causal effects of an intervention, as we illustrate below.

    Figure \ref{fig:toy_example} gives a toy example: the
    outcome $Y \in [0, 1]$, the
    probability of death, a binary treatment $A \in \{0, 1\}$ and a covariate
    $X \in \mathbb R$ which summarizes the patient health status (eg. the
    Charlson co-morbidity index \cite{charlson_new_1987}). We simulate a
    situation for which the treatment is beneficial (decreases
    mortality) for patients with high Charlson scores (bad health
    status). On the contrary, the treatment has little effect for patients in good
    condition (small Charlson scores).

    Figure \ref{fig:toy_example}a shows a random forest predictor with a
    counter-intuitive behavior: it predicts well on average the outcome (as
    measured by a regression $R^2$ score) but perform poorly to estimate
    causal quantities: the average treatment effect $\tau$ (as visible via
    the error $|\tau - \hat{\tau}|$) or the individual treatment
    effect (the error $\mathbb{E}[(\tau(x) - \hat{\tau}(x))^2]$).
    %
    On the contrary, Figure \ref{fig:toy_example}b shows a linear model with
    smaller $R^2$ score but better causal inference.%

    Intuitively, the problem is that causal estimation requires controlling
    an error on both treated and non-treated outcome for the same individual:
    the observed outcome, and the non-observed \emph{counterfactual} one.
    The linear model is misspecified --the outcome functions are not
    linear--, leading to poor $R^2$; but it interpolates better to regions
    where there are few untreated individuals --high Charlson score-- and
    thus gives better causal estimates. Conversely, the random forest puts
    weaker assumptions on the data, thus has higher $R^2$ score but is biased
    by the treated population in the region with poor overlap region, leading
    to bad causal estimates.

    This toy example illustrates that the classic minimum Mean Square Error
    criterion is not suited to choosing a model among a family of candidate
    estimators for causal inference.

@@

    A natural way to select a predictive model for causal inference would be
    an error measure between a causal quantity such as the conditional
    average treatment effects (CATE) and models' estimate. But such error is
    not a ``feasible'' risk: it cannot be computed solely from observed data
    and requires oracle knowledge.

@@

    Using eight simulations setups from \cite{powers_methods_2018}, where
    the oracle CATE is known, \citet{schuler_comparison_2018} compare four
    causal risks, concluding that for CATE estimation the best
    model-selection risk is the so-called $R\text{-risk}$
    \cite{nie_quasioracle_2017} --def.\,\ref{def:r_risk}, below. Their
    empirical results are clear for randomized treatment allocation but less
    convincing for observational settings where both simple Mean Squared
    Error --MSE, $\mu\text{-risk}(f)$ def.\,\ref{def:mu_risk}-- and
    reweighted MSE --$\mu\text{-risk}_{IPW}$ def.\,\ref{def:mu_ipw_risk}--
    appear to perform better than $R\text{-risk}$ on half of the simulations.
    Another work \cite{alaa_validating_2019} studied empirically both MSE and
    reweighted MSE risks on the semi-synthetic ACIC 2016 datasets
    \cite{dorie_automated_2019}, but did not include the $R\text{-risk}$ and
    looked only at the agreement of the best selected model with the true
    CATE risk --$\tau\text{-risk}(f)$ def.\, \ref{def:tau_risk}--, not on the
    full ranking of methods compared to the true CATE. We complete these
    prior empirical work by studying a wider variety of data generative
    processes and varying the influence of overlap, an important parameter of
    the data generation process which makes a given causal metric appropriate
    \cite{damour_overlap_2020}. We also study how to best adapt
    cross-validation procedures to causal metrics which themselves come with
    models to estimate.

@@

    Several theoretical works have proposed causal model selection procedures
    that are \emph{consistent}: select the best model in a family given
    asymptotically large data. These work typically rely on introducing a
    CATE estimator in the testing procedure. For instance matching
    \citep{rolling_model_2014}, an IPW estimate
    \citep{gutierrez_causal_2016}, a doubly robust estimator
    \citep{saito_counterfactual_2020}, or debiasing the error with influence
    functions \cite{alaa_validating_2019}. However, for theoretical
    guarantees to hold, the test-set correction needs to converge to the
    oracle: it needs to be flexible enough --or well-posed-- and asymptotic
    data. From a practical perspective, knowing that such requirements are
    met implies having a good CATE estimate, which amounts to having solved
    the original problem of causal model selection. We study how causal
    model-selection procedures behave outside of these settings.

@@

    Much work in causal inference has focused on building procedures that
    guarantee asymptotically consistent estimators. Targeted Machine Learning
    Estimation (TMLE) \cite{laan_targeted_2011,schuler_targeted_2017} and
    Double Machine Learning \cite{chernozhukov_double_2018} both provide
    estimators for Average Treatment Effect combining flexible treatment and
    outcome models. Here also, theories require asymptotic regimes and
    models to be \textit{well-specified}.

    By contrast, \citet{johansson_generalization_2021} studies causal
    estimation without assuming that estimators are well specified. They derive an
    upper bound on the oracle error to the CATE ($\tau\text{-risk}$) that involves
    the error on the outcome and the similarity of the distributions between the
    features of treated and control patients. However, they focus on using this
    upper bound for estimation, and do not give insights on model selection. In
    addition, for hyperparameter selection, they rely on a plugin estimate of the $\tau\text{-risk}$ built with
    counterfactual nearest neighbors, which has been shown ineffective
    \cite{schuler_comparison_2018}.

@@

    The Neyman-Rubin Potential Outcomes framework
    \cite{naimi2023defining,imbens_causal_2015} enables statistical
    reasoning on causal treatment effects: Given an outcome $Y \in \mathbb R$ (eg.
    mortality risk or hospitalization length), function of a binary treatment $A \in
        \mathcal{A} = \{0, 1\}$ (eg.~a medical act, a drug administration), and baseline
    covariates $X \in \mathcal{X} \subset \mathbb{R}^d$, we observe the factual
    distribution,
    $O = (Y(A), X, A) \sim \mathcal D = \mathbb P(y, x, a)$. However, we want to model the existence of potential observations
    (unobserved ie. counterfactual) that correspond to a different treatment. Thus we want
    quantities on the counterfactual distribution
    $O^{*} = (Y(1), Y(0), X, A) \sim \mathcal D^{*} = \mathbb P(y(1), y(0), x, a)$.

    Popular quantities of interest (estimands) are:
    at the population level, the
    Average Treatment Effect
    \begin{equation*}
        \text{(ATE)}\qquad
        \tau \myeq \; \mathbb{E}_{Y(1),Y(0) \sim \mathcal D^*}[Y(1) - Y(0)];
    \end{equation*}
    to model heterogeneity, the Conditional Average Treatment Effect
    \begin{equation*}
        \text{(CATE)}\qquad
        \tau (x) \myeq \; \mathbb{E}_{Y(1),Y(0) \sim \mathcal{D}^\star}[Y(1) - Y(0) | X=x].
    \end{equation*}

@@

    Some assumptions are necessary to assure identifiability of the causal estimands
    in observational settings \cite{rubin_causal_2005}. We assume the usual
    strong ignorability assumptions, composed of \emph{1)}
    \emph{unconfoundedness} $\{Y(0),
        Y(1) \} \indep A | X$, \emph{~2)} \emph{strong overlap} ie. every patient has a
    strictly positive probability to receive each treatment, \emph{3)}
    \emph{consistency}, and \emph{4)} \emph{generalization} (detailed in Appendix
    \ref{apd:causal_assumptions}). In this work, we investigate the
    role of the overlap \cite{damour_overlap_2020}, which is testable with
    data.

@@

    Should we know the two expected outcomes for a given $X$,
    we could compute the
    difference between them, which gives the causal effect of the treatment.
    %
    These two expected outcomes can be computed from the observed data:
    the consistency \ref{assumption:consistency} and unconfoundedness
    \ref{assumption:ignorability} assumptions imply the equality of two different
    expectations:
    \begin{equation}\label{eq:mu_identification}
        \mathbb E_{Y(a) \sim \mathcal{D^{\star}}} [Y(a)|X=x] = \mathbb E_{Y \sim \mathcal{D}} [Y|X=x, A=a]
    \end{equation}
    On the left, the expectation is taken on the counterfactual unobserved
    distribution. On the right, the expectation is taken on the factual observed
    distribution conditionally on the treatment. This equality is referred as the
    g-formula identification \cite{robins_new_1986}. For the rest of the
    paper, the expectations will always be taken on the factual observed
    distribution $\mathcal{D}$, and we will omit
    to explicitly specify the distribution. This identification leads to outcome based estimators (ie.
    g-computation estimators\cite{snowden_implementation_2011}), targeting the
    ATE $\tau$ with outcome modeling:
    \begin{equation}
        \tau = \mathbb E_{Y \sim \mathcal{D^{\star}}}[Y(1) - Y(0)|X=x] = \mathbb E_{Y \sim \mathcal{D}}[Y|A=1] - \mathbb E_{Y \sim \mathcal{D}}[Y| A=0]
        \label{eq:tau_population}
    \end{equation}
    This equation has two central quantities: the conditional expectancy
    function of the outcome associated to specific covariates and
    treatment or not is, often called the \emph{response function}:
    \begin{equation}
        \text{(Response function)}
        \qquad
        \mu_{a}(x) \myeq \; \mathbb E_{Y \sim \mathcal{D}} [Y|X=x, A=a].
        \label{def:mu_a}
    \end{equation}

@@

    Similarly to the ATE, for the CATE, at the individual level:
    \begin{equation}
        \tau(x) = \mu_{1}(x) - \mu_{0}(x)
        \label{eq:cate_estimate}
    \end{equation}

    \paragraph{Robinson decomposition}
    Another decomposition of the outcome model plays in important role,
    the R-decomposition
    \cite{robinson_rootnconsistent_1988}:
    introducing two quantities, the conditional mean outcome
    and the probability to be treated (known as propensity score \cite{rosenbaum_central_1983}):
    \begin{align}
        \text{(Conditional mean outcome)} \qquad  m(x) \myeq \; & \mathbb E_{Y \sim \mathcal{D}} [Y|X=x].
        \label{def:m}
        \\
        \text{(Propensity score)}\qquad
        e(x) \myeq \;                                           & \mathbb P[A=1|X=x],
        \label{def:propensity_score}
    \end{align}
    the outcome can be written
    \begin{equation}\label{eq:r_decomposition}
        \text{(R-decomposition)}\qquad y(a) = m(x) + \big( a - e(x) \big) \tau(x) + \varepsilon(x; a) \quad\text{with}\quad \mathbb E[\varepsilon(X; A)|X, A] = 0
    \end{equation}
    $m$ and $e$ are often called
    \emph{nuisances} \cite{chernozhukov_double_2018}; they are unknown in
    general.

    %As noted by \cite{johansson_generalization_2021}, the machine learning
    %community often referred to the CATE by ITE, the Individual Treatment Effect.
    %From a purely causal point of view, the ITE is uniquely defined for each
    %individual and might not be accessible: $ITE(x_i) = Y_i(1) -  Y_i(0)$. On the
    %contrary, the CATE can always be derived by taking conditional expectancies. It
    %is the expected effect of the treatment in the region of the covariate space
    %around X. %Too cultivate

@@

    \begin{figure}[h!]
        \centering
        \includegraphics[width=0.7\linewidth]{images/estimation_procedure_causal_selection_procedure.png}
        \caption{Estimation procedure for causal model
            selection.}\label{problem:estimation_procedure:figure}
    \end{figure}

@@

    \paragraph{Data Generation Process}

>>>>

%\documentclass[MYBIB,STIX1COL]{WileyNJD-v2}
\documentclass[10pt,twocolumn]{article}

\usepackage[a4paper, left=15mm, right=15mm, top=20mm,]{geometry}
% declared functions and additional packages
\usepackage{caption}
\usepackage{subcaption}
\usepackage{threeparttable}
\usepackage{times}
\usepackage{hyperref}
\def\mycitecolor{green!50!black}
\def\mylinkcolor{red!60!black}
\hypersetup{ colorlinks=true, linkcolor=\mylinkcolor, citecolor=\mycitecolor }
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{amsmath}
\usepackage{amssymb} %mathbb packages
\usepackage{siunitx}
\usepackage{algorithm, algorithmicx, algpseudocode}
\usepackage{multirow}
\usepackage[sorting=nyt,style=numeric, natbib=true,doi=false,isbn=false,url=false,eprint=false, maxnames=999, maxcitenames=2]{biblatex} 
\DeclareNameAlias{default}{last-first}
\usepackage{placeins}
% associated arxiv class
%\usepackage{arxiv}
\usepackage{multicol}
\setlength{\columnsep}{0.5cm}
\usepackage{wrapfig}
%\usepackage{wraptable}
\usepackage{authblk}% for author affiliations
\usepackage[dvipsnames]{xcolor}

@@

\setlength{\marginparwidth}{0.5cm}
%\setlength\textwidth{177.8mm}

@@

\newtheorem{definition}{Definition}
\newtheorem{condition}{Condition}
\newtheorem{property}{Property}
\newtheorem{remark}{Remark}
\newtheorem{notation}{Notation}
\newtheorem{proposition*}{Proposition}
\newtheorem{proposition}{Proposition}
\newtheorem{lemma}{Lemma}
\newtheorem{theorem}{Theorem}
\newtheorem{assumption}{Assumption}
\newtheorem{proof}{Proof}
%
\def\keywordname{{\bfseries \emph{Keywords}}}%
\def\keywords#1{\par\addvspace\medskipamount{\rightskip=0pt plus1cm
\def\and{\ifhmode\unskip\nobreak\fi\ $\cdot$
}\noindent\keywordname\enspace\ignorespaces#1\par}}

@@

\affil[1]{Inria, Soda, Saclay, France}
\affil[2]{Mission Data, Haute Autorité de Santé, Saint-Denis, France}
\affil[*]{Corresponding author: m.doutreligne@has-sante.fr}

@@

\onecolumn
\maketitle
\begin{abstract}
    \textbf{Background:}
    As predictive models --\emph{eg} from machine learning-- give likely
    outcomes, they may be used to reason on the effect of an intervention,
    a causal-inference task. The increasing complexity of
    health data has opened the door to a plethora of models, but also the
    Pandora box of model selection: which of these models yield the most
    valid causal estimates? Causal model selection should control both outcome errors for each
    individual, treated or not treated, whereas only one outcome is observed.
    Theoretically, simple risks used in machine learning do not control causal
    effects when treated and non-treated population differ too much. \\
    \textbf{Methods:}
    More elaborate risks build proxies of the causal error using ``nuisance''
    re-weighting to compute it on the observed data. But does computing these
    nuisance adds noise to model selection? Drawing from an extensive empirical
    study based on a simulation and three semi-simulated datasets based on real
    covariates, we compare different metrics for causal model selection.\\
    \textbf{Results:} Among all metrics the classical mean squared error (ie.
    factual $\mu\mathrm{-risk}$) is worse and re-weighting it with propensity score
    ($\mu\mathrm{-risk}_{IPW}$) does not bring much improvements. The
    $R\text{-risk}$, which includes a model of mean outcome and propensity
    scores, leads to the best performances.\\
    \textbf{Conclusions:} We outline a good causal model-selection procedure
    using the so-called $R\text{-risk}$; using flexible estimators to compute
    the nuisance models on the train set; and splitting out 10\% of the data to
    compute risks.
\end{abstract}
\keywords{Model Selection, Treatment Effect, G-formula, Observational Study, Machine Learning, }
\clearpage

\twocolumn
\section{Background}\label{sec:intro}

@@

\begin{figure}[t!]
    %\centering
    {\sffamily\footnotesize\scalebox{1.03}{a) Random forest, good average prediction but bad
            causal inference}}

    \includegraphics[width=1\linewidth]{images/toy_random_forest_high_R2_high_tau_risk.pdf}%

    {\sffamily\footnotesize\scalebox{1.03}{b) Linear model, worse average
            prediction but better causal inference}}

    \hfill%
    \includegraphics[width=1\linewidth]{images/toy_tlinear_model_small_R2_small_tau_risk.pdf}%

    \caption[The best predictor may not estimate best causal
        effects]{\textbf{Illustration}: a) a random-forest estimator
        with high performance for standard prediction (high $R^2$) but that
        yields poor causal estimates (large error between true effect $\tau$ and
        estimated $\hat{\tau}$), b) a linear estimator with smaller
        prediction performance leading to better causal estimation. \\[1ex]
        Selecting the estimator with the smallest error to the individual
        treatment effect $\mathbb{E}[(\tau(x) - \hat{\tau}(x))^2]$
        --the $\tau\text{-risk}$, def.\,\ref{def:tau_risk} -- would lead to
        the best causal estimates; however computing this error is not
        feasible: it requires access to unknown quantities:
        $\tau(x)$. \\[1ex]
        While the random forest fits the data better than the linear model, it
        gives worse causal inference because its error is inhomogeneous between
        treated and untreated. The $R^2$ score does not capture this
        inhomogeneity.
        \label{fig:toy_example}
    }
\end{figure}

Using a predictor to reason on causal effects relies on contrasting the
prediction of the outcome for a given individual with and without the
treatment --as detailed in \autoref{sec:neyman_rubin}.
%
Given various predictors of the outcome, which one should we use?
%
Standard predictive modeling or machine-learning practice selects the
predictor that minimizes the expected error.
However, this predictor may not be the best model to reason about
causal effects of an intervention, as we illustrate below.

Figure \ref{fig:toy_example} gives a toy example: the probability $Y$ of
an undesirable
outcome (\emph{eg} death), a binary treatment $A \in \{0, 1\}$, and a covariate
$X \in \mathbb R$ summarizing the patient health status (eg. the
Charlson index \cite{charlson_new_1987}). We simulate a
treatment beneficial (decreases $Y$) for patients with high Charlson scores (bad health
status) but with little effect for patients in good
condition (low Charlson scores).

Figure \ref{fig:toy_example}a shows a random forest predictor with a
counter-intuitive behavior: it predicts well on average the outcome (as measured
by a regression $R^2$ score) but perform poorly to estimate causal quantities:
the average treatment effect $\tau$ (as visible via the error $|\tau -
    \hat{\tau}|$) or the conditional average treatment effect (the error
$\mathbb{E}[(\tau(x) - \hat{\tau}(x))^2]$, called CATE).
%
On the contrary, Figure \ref{fig:toy_example}b shows a linear model with
smaller $R^2$ score but better causal inference.%

The problem is that causal estimation requires controlling
an error on both treated and non-treated outcome for the same individual:
the observed outcome, and the non-observed \emph{counterfactual} one.
The linear model is misspecified --the outcome functions are not
linear--, leading to poor $R^2$; but it interpolates better to regions
where there are few untreated individuals --high Charlson score-- and
thus gives better causal estimates. Conversely, the random forest puts
weaker assumptions on the data, thus has higher $R^2$ score but is biased
by the treated population in the poor-overlap region, leading
to bad causal estimates.

This toy example illustrates that the classic minimum Mean Square Error
criterion is not suited to choosing a model among candidate
estimators for causal inference.

@@

A natural way to select a predictive model for causal inference would be
an error measure between a causal quantity such as the CATE and models' estimate. But such error is
not a ``feasible'' risk: it cannot be computed solely from observed data
and requires oracle knowledge.

@@

Using eight simulations setups from \cite{powers_methods_2018}, where
the oracle CATE is known, \citet{schuler_comparison_2018} compare four
causal risks, concluding that for CATE estimation the best
model-selection risk is the so-called $R\text{-risk}$
\cite{nie_quasioracle_2017} --def.\,\ref{def:r_risk}, below. Their
empirical results are clear for randomized treatment allocation but less
convincing for observational settings where both simple Mean Squared
Error --MSE, $\mu\text{-risk}(f)$ def.\,\ref{def:mu_risk}-- and
reweighted MSE --$\mu\text{-risk}_{IPW}$ def.\,\ref{def:mu_ipw_risk}--
appear to perform better than $R\text{-risk}$ on half of the simulations.
Another work \cite{alaa_validating_2019} studied empirically both MSE and
reweighted MSE risks on the semi-synthetic ACIC 2016 datasets
\cite{dorie_automated_2019}, but did not include the $R\text{-risk}$. We complete these
prior empirical work by studying a wider variety of data generative
processes and varying the influence of overlap, an important parameter of
the data generation process which makes a given causal metric appropriate
\cite{damour_overlap_2020}. We also study how to best adapt
cross-validation procedures to causal metrics which themselves come with
models to estimate.

@@

Several theoretical works have proposed causal model selection procedures
that are \emph{consistent}: select the best model in a family given
asymptotically large data. These work rely on introducing a
CATE estimator in the testing procedure: matching
\citep{rolling_model_2014}, an IPW estimate
\citep{gutierrez_causal_2016}, a doubly robust estimator
\citep{saito_counterfactual_2020}, or debiasing the error with influence
functions \cite{alaa_validating_2019}. However, for theoretical
guarantees to hold, the test-set correction needs to converge to the
oracle: it needs to be flexible enough --well-posed-- and asymptotic
data. From a practical perspective, meeting such requirements
implies having a good CATE estimate, thus having solved
the original problem of causal model selection.

@@

Much work in causal inference has focused on procedures that
guarantee asymptotically consistent estimators, such as Targeted
Machine Learning
Estimation (TMLE) \cite{laan_targeted_2011,schuler_targeted_2017} or
Double Machine Learning \cite{chernozhukov_double_2018}. Here also, theories require asymptotic regimes and
models to be \textit{well-specified}.

By contrast, \citet{johansson2022generalization} studies causal estimation
without assuming that estimators are well specified. They derive an upper bound
on the oracle error to the CATE ($\tau\text{-risk}$) that involves the error on
the outcome and the similarity of the distributions of treated and control
patients. However, they use this upper bound for model optimization,
and do not give insights on model selection. In addition, for hyperparameter
selection, they rely on a plugin estimate of the $\tau\text{-risk}$ built with
counterfactual nearest neighbors, which has been shown ineffective
\cite{schuler_comparison_2018}.

@@

The Neyman-Rubin Potential Outcomes framework
\cite{naimi2023defining,imbens_causal_2015} enables statistical reasoning on
causal treatment effects: Given an outcome $Y \in \mathbb R$ (eg. mortality risk
or hospitalization length), function of a binary treatment $A \in \mathcal{A} =
    \{0, 1\}$ (eg.~a medical procedure, a drug administration), and baseline
covariates $X \in \mathcal{X} \subset \mathbb{R}^d$, we observe the factual
distribution, $O = (Y(A), X, A) \sim \mathcal D = \mathbb P(y, x, a)$. However,
we want to model the existence of potential observations (unobserved ie.
counterfactual) that correspond to a different treatment. Thus we want
quantities on the counterfactual distribution $O^{*} = (Y(1), Y(0), X, A) \sim
    \mathcal D^{*} = \mathbb P(y(1), y(0), x, a)$.

Popular quantities of interest (estimands) are:
at the population level, the
Average Treatment Effect
\begin{flalign*}
    \text{ATE} &  &
    \tau \myeq \; \mathbb{E}_{Y(1),Y(0) \sim \mathcal D^*}[Y(1) - Y(0)];
               &  &
\end{flalign*}
at the individual level, to model heterogeneity, the Conditional Average Treatment Effect
\begin{flalign*}
    \text{CATE} &  &
    \tau (x) \myeq \; \mathbb{E}_{Y(1),Y(0) \sim \mathcal{D}^\star}[Y(1) - Y(0) | X=x].
                &  &
\end{flalign*}

@@

Assumptions are necessary for causal estimands to be
identifiability
in observational settings \cite{rubin_causal_2005}. We assume the usual
strong ignorability assumptions: \emph{1)}
\emph{unconfoundedness} \mbox{$\{Y(0),
        Y(1) \} \indep A | X$}, \emph{~2)} \emph{strong overlap} ie. every patient has a
strictly positive probability to receive each treatment, \emph{3)}
\emph{consistency}, and \emph{4)} \emph{generalization} (detailed in
\ref{apd:causal_assumptions}).

@@

Should we know the two expected outcomes for a given $X$,
we could compute the
difference between them, which gives the causal effect of the treatment.
%
These two expected outcomes can be computed from the observed data:
the consistency \ref{assumption:consistency} and unconfoundedness
\ref{assumption:ignorability} assumptions imply the equality of two different
expectations:
\begin{equation}\label{eq:mu_identification}
    \mathbb E_{Y(a) \sim \mathcal{D^{\star}}} [Y(a)|X=x] = \mathbb E_{Y \sim \mathcal{D}} [Y|X=x, A=a]
\end{equation}
On the left, the expectation is taken on the counterfactual unobserved
distribution. On the right, the expectation is taken on the factual observed
distribution conditionally on the treatment. This equality is referred as the
g-formula identification \cite{robins_new_1986}. For the rest of the
paper, the expectations will always be taken on the factual observed
distribution $\mathcal{D}$. This identification leads to outcome based estimators (ie.
g-computation estimators \cite{snowden_implementation_2011}), targeting the
ATE $\tau$ with outcome modeling:
\begin{eqnarray}
    \tau =& \mathbb E_{Y \sim \mathcal{D^{\star}}}[Y(1) - Y(0)|X=x]
    \notag
    \\
    = &\mathbb E_{Y \sim \mathcal{D}}[Y|A=1] - \mathbb E_{Y \sim \mathcal{D}}[Y| A=0]
    \label{eq:tau_population}
\end{eqnarray}
This equation builds on two quantities: the conditional expectancy
of the outcome given the covariates and either
treatment or no no treatment, called \emph{response function}:
\begin{flalign*}
    \text{Response function}
     &  &
    \mu_{a}(x) \myeq \; \mathbb E_{Y \sim \mathcal{D}} [Y|X=x, A=a]
     &  &
\end{flalign*}

@@

Similarly to the ATE, at the individual level, the CATE:
\begin{equation}
    \tau(x) = \mu_{1}(x) - \mu_{0}(x)
    \label{eq:cate_estimate}
\end{equation}

\paragraph{Robinson decomposition}
The \emph{R-decomposition}
of the outcome model plays an important role,
\cite{robinson_rootnconsistent_1988}:
introducing two quantities, the conditional mean outcome
and the probability to be treated (known as propensity score \cite{rosenbaum_central_1983}):
\begin{flalign}
    \text{Conditional mean outcome} &                    & m(x) \myeq \; & \mathbb E_{Y \sim
    \mathcal{D}} [Y|X=x]            &                    &
    \label{def:m}
    \\
    \text{Propensity score}         &                    &
    e(x) \myeq \;                   & \mathbb P[A=1|X=x]
    \label{def:propensity_score}
\end{flalign}
the outcome can be written
\begin{flalign}\label{eq:r_decomposition}
    \text{R-decomposition}            &  & y(a) = m(x) + \big( a - e(x) \big)
    \tau(x) + \varepsilon(x; a)\notag &  &
    \\\ &&\text{with}\quad \mathbb E[\varepsilon(X; A)|X, A] = 0&&
\end{flalign}
$m$ and $e$ are often called
\emph{nuisances} \cite{chernozhukov_double_2018}; they are unknown.

%As noted by \cite{johansson2022generalization}, the machine learning
%community often referred to the CATE by ITE, the Individual Treatment Effect.
%From a purely causal point of view, the ITE is uniquely defined for each
%individual and might not be accessible: $ITE(x_i) = Y_i(1) -  Y_i(0)$. On the
%contrary, the CATE can always be derived by taking conditional expectancies. It
%is the expected effect of the treatment in the region of the covariate space
%around X. %Too cultivate

\begin{table*}[!tb]
    \makebox[\textwidth]{
        \begin{threeparttable}[b]
            \caption{Review of causal risks
                ---
                The $R\text{-risk}^*$ is called $\tau \text{-risk}_R$ in
                \citet{schuler_comparison_2018}.
                \label{tab:evaluation_metrics}}
            \centering
            \begin{tabular}{llr}
                \toprule
                Risk                                                                                           & Equation
                                                                                                               & Reference                                                                                                                                           \\
                \midrule
                $mse(\tau(X), \tau_f(X))=\tau\text{-risk}$                                                     & $\mathbb E_{X\sim
                            p(X)}[(\tau(X) - \hat \tau_f(X))^2] $
                                                                                                               & Eq. \ref{eq:tau_risk} \cite{hill_bayesian_2011}                                                                                                     \\
                $mse(Y, f(X)) = \mu\text{-risk}$                                                               & $\mathbb{E}_{(Y, X, A)
                        \sim \mathcal D}\left[(Y-f(X ; A))^2 \right]$
                                                                                                               & Def. \ref{def:mu_risk} \cite{schuler_comparison_2018}                                                                                               \\
                $\mu\text{-risk}_{IPW}^*$                                                                      & $\mathbb{E}_{(Y, X, A)
                        \sim \mathcal D}\left[ \Big( \frac{A}{e(X)} + \frac{1-A}{1-e(X)} \Big)
                (Y-f(X ; A))^2 \right]$                                                                        & Def.
                \ref{def:mu_ipw_risk} \cite{vanderlaan_unified_2003}                                                                                                                                                                                                 \\
                $\tau\text{-risk}^{\star}_{IPW}$                                                               & $\mathbb{E}_{(Y, X, A) \sim \mathcal D} \left[ \Big(Y \left( \frac{A}{e(X)} - \frac{1-A}{1-e(X)}\right)-\hat \tau_f\left(X\right)\Big)^2 \right]$ &
                Def. \ref{def:tau_ipw_risk} \cite{wager_estimation_2018}
                \\
                $U\text{-risk}^*$                                                                              & $\mathbb{E}_{(Y, X, A) \sim \mathcal D}  \big[
                \big( \frac{Y-m\left(X\right)}{A-e\left(X\right)} -  \hat \tau_f\left(X\right)\big)^{2} \big]$ &
                Def. \ref{def:u_risk} \cite{nie_quasioracle_2017}
                \\
                $R\text{-risk}^*$                                                                              & $\mathbb{E}_{(Y, X, A)
                        \sim \mathcal D} \big[\big(\left(Y-m\left(X\right)\right)
                -\left(A-e\left(X\right)\right) \hat \tau_f\left(X\right)\big)^{2} \big]$                      &
                Def. \ref{def:r_risk} \cite{nie_quasioracle_2017}
                \\
                \bottomrule
            \end{tabular}
        \end{threeparttable}
    }
\end{table*}

@@

\begin{figure}[h!]
    \includegraphics[width=\linewidth]{images/estimation_procedure_causal_selection_procedure.png}
    \caption{Estimation procedure for causal model
        selection.}\label{problem:estimation_procedure:figure}
\end{figure}

@@

\paragraph{Data Generation}
